A thermodynamic view of dusty protoplanetary disks

### 2017-08-08 linear.tex
final revision before submission, spell-checked

\section{Application to protoplanetary disks}\label{linear_problem}
We now examine instabilities in dusty protoplanetary disks
based on explicit descriptions of the equilibrium state. We first
specify explicit disk structures to analyze. We show 
that sharp radial edges in the dust component are unstable. We then
revisit two well-known instabilities in dusty gas using the
hydrodynamic framework developed thusfar, namly secular graviational
instabilities and the streaming instability.

@@

\subsubsection{Vertical shear}\label{vertshear}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp})
gives the disk's explicit vertical shear profile:
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

@@

\subsubsection{Dusty vertical buoyancy}\label{vbuoyancy}
For the above equilibria the vertical buoyancy frequency is given
explicitly as 
\begin{align}
  N_z^2 =
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where we have used $c_s=c_s(r)$. Then 
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

\subsection{Instability of dusty edges} 
Here we apply the dusty analog of the Solberg-Hoiland criteria derived
in \S\ref{iso_perfect} assuming strictly isothermal gas. As discussed
there, the first criterion in generally satisifed. Thus we only
consider the second condition, Eq. \ref{dusty_solberg2}.

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

@@

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
% \chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
%   \left[
%   \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
%   \sqrt{
%   \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
%   4 \frac{\Htilde^2}{H_\mathrm{g}^2}
%   \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
%   }
%   \,\right]. 
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{P}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{P}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_rP<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled. 
%(So the dust layer thickness $\Hd\gg 2
%\times10^{-3}\Hgas$ as well.) 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

@@

\subsection{Radially localized linear problem}
We now specialize further and compute explicit solutions to the linear
problem. We consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta X (r, z) = \delta X_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd X_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for 
vertically isothermal dusty gas with the pressure
equation in place of the dust-fraction
(Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla
P/\rho$. We have temporarily restored self-gravity  to
discuss SGI in the next
section. The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
%Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_sg} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We next discuss solutions to these equations.  We first
show that the above equations yield the secular
gravitational instability and streaming instability in the strong
drag regime in \S\ref{sgi} and \S\ref{si}, respectively. We then
consider 3D, stratified disks in \S\ref{results} to study the dusty
vertical shear instability.

@@

\subsubsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi(z=0) = -2\pi G
\Sigma W/\left|k_x\right|$.%, where $\Sigma$ is the total surface
%density.

@@

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi&\equiv \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta
  P\Delta\rho^*\right).\notag 
%\notag\\
%       &= \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta\rhog^*\Delta\rhod\right).\notag
\end{align*}
(Note that it is important to include 
the global radial pressure gradient in $\Delta P = \delta P + \ii \dd v_x \p_rP/\sigma$.) Then 
$\varphi > 0 $ indicates gas pressure lagging behind total density, 
which is true for all the cases. Thus SI is indeed associated with
such a phase lag.

@@

In Fig. \ref{si_compare_fig} we calculate the most unstable SI 
mode as a function of $\taus$ at fixed $K_z=30$ and $\epsilon =3$. We
compare results between the full, two-fluid linear analysis and the
one-fluid framework. We also include an analytic model, described in Appendix 
\ref{si_dust_rich}, based on the one-fluid dispersion relation with 
additional approximations (orange diamonds).

Both results based on the one-fluid framework compares well with the 
full, two-fluid analysis, only breaking down at a relatively large
$\taus\OmK\gtrsim 0.1$. For larger $\taus$ the two-fluid phase
lag drops, along with the growth rate. This suggest that a
non-vanishing phase lag is necessary for instability. However, the
magnitude of the phase lag does not correlate with growth rates. In
fact, $\varphi$ remains finite as $|\sigma|, \taus\to0$ (but
non-zero). This arises because the optimum $K_x\propto \taus^{-1/2}$
diverges (see Appendix \ref{si_dust_rich}).

>>>>

\section{Application to protoplanetary disks}\label{linear_problem}
We now examine instabilities in dusty protoplanetary disks
based on explicit descriptions of the equilibrium state. We first
specify the disk structures to be analyzed. We show  
that sharp radial edges in the dust-to-gas ratio can render the disk unstable. We then 
revisit two well-known instabilities in dusty gas using the
hydrodynamic framework developed thus far, namely secular gravitational (SGI) 
instabilities and the streaming instability (SI).

@@

\subsubsection{Vertical shear}\label{vertshear}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp})
gives the disk's explicit vertical shear profile:
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, which survives in the dust-free limit.

@@

\subsubsection{Dusty vertical buoyancy}\label{vbuoyancy}
For the above equilibrium the vertical buoyancy frequency is given
explicitly as 
\begin{align}
  N_z^2 =
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where we have used $c_s=c_s(r)$. Then 
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

\subsection{Instability of dusty edges} 
Here we apply the dusty analog of the Solberg-Hoiland criteria derived
in \S\ref{iso_perfect} assuming strictly isothermal gas. As discussed
there, the first criterion is generally satisfied. Thus we only
consider the second condition, Eq. \ref{dusty_solberg2}.

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occurred). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

@@

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
left hand side is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
% \chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
%   \left[
%   \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
%   \sqrt{
%   \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
%   4 \frac{\Htilde^2}{H_\mathrm{g}^2}
%   \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
%   }
%   \,\right]. 
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{P}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{P}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_rP<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled. 
%(So the dust layer thickness $\Hd\gg 2
%\times10^{-3}\Hgas$ as well.) 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short radial lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

@@

\subsection{Radially local problem}
We now specialize further and compute explicit solutions to the linear
problem. We consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta X (r, z) = \delta X_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd X_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for 
vertically isothermal dusty gas with the pressure
equation in place of the dust-fraction
(Eq. \ref{masseq}, \ref{momeq}, \ref{poisson}, \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla
P/\rho$. We have temporarily restored self-gravity  to
discuss SGI in the next
section. The linearized dust-diffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
%Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_sg} is a set of ordinary
differential equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We next discuss solutions to these equations.  We first
show that the above equations yield the SGI and SI in the strong
drag regime in \S\ref{sgi} and \S\ref{si}, respectively. We then
consider 3D, stratified disks in \S\ref{results} to study how the addition of  dust 
affects the vertical shear instability.

@@

\subsubsection{Secular gravitational instability}\label{sgi}
Consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$, where $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Here $\epsilon = \Sigma_\mathrm{d}/\Sigma_\mathrm{g}$ and 
$\tepsilon=\Sigma_\mathrm{d}/\Sigma$, where $\Sigma_\mathrm{d,g}$ are the dust and gas surface densities. 
The background disk is 
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi(z=0) = -2\pi G
\Sigma W/\left|k_x\right|$.%, where $\Sigma$ is the total surface
%density.

@@

In Table \ref{si_compare} we also calculate the phase difference
between the Lagrangian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi&\equiv \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta
  P\Delta\rho^*\right).\notag 
%\notag\\
%       &= \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta\rhog^*\Delta\rhod\right).\notag
\end{align*}
(Note that it is important to include 
the global radial pressure gradient in $\Delta P = \delta P + \ii \dd v_x \p_rP/\sigma$.) Then 
$\varphi > 0 $ indicates gas pressure lagging behind total density, 
which is true for all the cases. Thus SI is indeed associated with
such a phase lag.

@@

In Fig. \ref{si_compare_fig} we calculate the most unstable SI 
mode as a function of $\taus$ at fixed $K_z=30$ and $\epsilon =3$. Growth rates are maximized over $K_x$. 
We 
compare results between the full, two-fluid linear analysis and the
one-fluid framework. We also include an analytic model, developed in Appendix 
\ref{si_dust_rich}, based on the one-fluid dispersion relation with 
additional approximations (orange diamonds).

Both results based on the one-fluid framework compares well with the 
full, two-fluid analysis, only breaking down at a relatively large
$\taus\OmK\gtrsim 0.1$. For larger $\taus$ the two-fluid phase
lag drops, along with the growth rate. This suggest that a
non-vanishing phase lag is indeed necessary for instability. However, the
magnitude of the phase lag does not correlate with growth rates. In
fact, $\varphi$ remains finite as $|\sigma|, \taus\to0$ (but
non-zero). This arises because the optimum $K_x\propto \taus^{-1/2}$
diverges (see Appendix \ref{si_dust_rich}).
